An introduction to Docker for reproducible research, with examples from the R environment

### 2014-09-16 manuscript.Rmd
getting there. more writing

Docker is an open source (Apache Version 2) software platform written in
the Go programming language.

Docker builds on many long-familiar technologies from operating
systems research: LXC containers, virtualization of the OS, and
a hash-based or git-like verioning and differencing system, among
others. Researchers familiar with LXC may be tempted to view the Docker
system as little more than a creative application and marketing of an
existing technology. However, the Docker approach adds several features
crucial for its use in reproducible research (adapted from [Docker
FAQ](https://docs.docker.com/faq/#what-does-docker-add-to-just-plain-lxc)):

1. **"Dependency Hell"**
2.  **Imprecise documentation**
3.  **Code rot**
4. **Barriers to adoption and reuse in existing solutions**

------

# Using Docker in an R environment

@@

@@

In the R environment, the open source RStudio IDE provides
another key component in making this system accessible to most
domain scientists. Because the RStudio IDE is written completely with
standard web languages: javascript, css, and html, its web-based client,
RStudio-Server, looks and feels identical to its popular desktop IDE.
RStudio Server provides a way to interact with R on a remote environment
without the latency, X tunnelling, etc typically involved in running R
on a remote server.

@@

- Interactive use, scripted use (Dockerfiles)
- Reverting to a previous state
- Integration with local development environment
- Integration with cloud computing
- Transparent & extensible
- serial and linking containers

- Best practices are still taking shape

## Docker remix

@@

<!-- excerpted -->

1) Remix.  Titus has an excellent post, "Virtual Machines Considered
Harmful for reproducibility" [1] , .  In contrast,
Docker containers are designed to work exactly like that -- reusable
building blocks you can link together with very little overhead in
disk space or computation.  This more than anything else sets Docker
apart from the standard VM approach.

Docker is not a virtual machine; containers are designed expressly to
be remixable blocks.  You can put an R engine on one container and a
mysql database on another and connect them. Docker philosophy aims at
one function per container to maximize this reuse.  of course it's up
to you to build this way rather than a single monolithic dockerfile,
but the idea of linking containers is a technical concept at the heart
of docker that offers a second and very different way to address the
'remix' problem of VMs.

 Provisioning scripts. Docker images are not 'black boxes'. A
"Dockerfile" is a simple make-like script which installs all the
software necessary to re-create ("provision") the image.  This has
many advantages: (a) The script is much smaller and more portable than
the image. (b) the script can be version managed (c) the script gives
a human readable (instead of binary) description of what software is
installed and how. This also avoids pitfalls of traditional
documentation of dependencies that may be too vague or out-of-sync.
(d) Other users can build on, modify, or extend the script for their
own needs.  All of this is what we call the he "DevOpts" approach to
provisioning, and can be done with AMIs or other virtual machines
using tools like Ansible, Chef, or Puppet coupled with things like
Packer or Vagrant (or clever use of make and shell scripts).

For a much better overview of this "DevOpts" approach in the
reproducible research context and a gentle introduction to these
tools, I highly recommend taking a look at Clark et al [2].

@@

Related challenges:  RStudio and a consistent environment. 
- Linked volumes 
- linked containers 
- signed containers?

@@

>>>>

<!-- Docker is an open source (Apache Version 2) software platform written in
the Go programming language.-->

Docker is an open source project that builds on many long-familiar
technologies from operating systems research: LXC containers,
virtualization of the OS, and a hash-based or git-like verioning
and differencing system, among others.

<!-- Define:
- Dockerfile
- Image 
- Container: When we run a docker image, we get a container.  We can run the same image multiple times, resulting in many identical containers.  We can save the state of a container at any time, which creates a new image.  
- Dockerfile -> build -> image
- Image -> run -> container
-->

I introduce the most relevant concepts from Docker through the 
context of the four challenges for reproducible research I have
discussed above.

### 1. Docker images: resolving 'Dependency Hell'

A Docker based approach works similarly to a virtual machine image in
addressing the dependency problem by providing other researchers with
a binary image in which all the software has already been installed,
configured and tested. (A machine image can also include all data files
necessary for the research, which may simplify the distribution of data.)

A key difference between Docker images and other virutal machines is 
that the Docker images share the Linux kernel with the host machine.
For the end user the primary consequence of this is that any Docker
image must be based on a Linux system with Linux-compatible software, which
includes (R, Python, Matlab, and most other scientific programming needs). [^4]

Sharing the Linux kernel makes Docker much more light-weight and higher
performance than complete virtual machines -- a typical desktop computer
could run no more than a few virutal machines at once but 
would have no trouble running 100s of Docker containers (a container
is simply the term for running instance of an image). This feature
has made Docker particularly attractive to industry and is largely
responsible for the immense popularity of Docker. For our purposes this
is a nice bonus, but the chief value to reproducible research lies
in other aspects.

[^4]: Note that re-distribution of an image in which proprietary software
has been installed will be subject to any relevant licensing agreement.

<!-- 
- image sharing (hub)
- image linking
-->

### 2. Dockerfiles: Resolving imprecise documentation

Though Docker images can be created interactively, this leaves little
transparent record [^3] of what software has been installed and how.
Dockerfiles provide a simple script (similar to a Makefile) that defines
exactly how to build up the image, consistent with the DevOpts approach
we mentioned previously.

With a syntax that is simpler than other provisioning tools (e.g. Chef,
Puppet, Ansible) or Continuous Integration (CI) platforms (e.g. Travis CI,
Shippable CI); users need little more than a basic familarity with shell
scripts and a Linux distribution software environment (e.g. Debian-based
`apt-get`) to get started writing Dockerfiles.

This approach has many advantages:

- While machine images can be very large (many gigabytes), a Dockerfile
is just a small plain text file that can be easily stored and shared.

- Small plain text files are ideally suited for use with a version
management system such as `subversion` or `git`, which can track any
changes made to the `Dockerfile`

- the `Dockerfile` provides a human readable summary of the necessary
software dependencies, environmental variables and so forth needed to
execute the code. There is little possiblity of the kind of holes or 
imprecision in such a script that so frequently cause difficulty in
manually implemented documentation of dependencies. This approach
also avoids the burden of having to tediously document dependencies 
at the end of a project, since they are instead documented as they 
are installed by writing the `Dockerfile`.

- Unlike a `Makefile` or other script, the `Dockerfile` includes all
software dependencies down to the level of the OS, and is built by the
Docker `build` tool, making it very unlikely that the resulting build
will differ when being built on different machines. This is not to say
that all builds of a Dockerfile are bitwise identical.  In particular,
builds executed later will install more recent versions of the same
software, if available, unless the package managers used are explicitly
configured otherwise. We address this issue in the next section.

- It is possible to add checks and tests following the commands for
installing the software environment, which will verify that the setup
has been successful.  This can be important in addressing the issue of
code-rot which we discuss next.

- It is straight-straight forward for other users to extend or customize 
the resulting image by editing the script directly.

[^3]: The situation is in fact slightly better than the virtual machine
approach because these changes are versioned. Docker provides tools
to inspect differences (diffs) between the images, and we can also roll
back changes to earlier versions.

### 3. Tackling code-rot with image versions

As I have discussed above, changes to the dependencies, whether they are
the result of security fixes, new features, or deprecation of old software,
can break otherwise functioning code. These challenges can be significantly
reduced because Docker defines the software environment to a particular operating system 
and suite of libraries, such as the Ubuntu or Debian distribution.  Such 
distributions use a staged release model with `stable`, `testing` and `unstable` 
phases subjected to extensive testing to catch such potential problems [Ooms 2013],
while also providing regular security updates to software within each stage.
Nonetheless, this cannot completely avoid the challenge of code-rot,
particularly when it is necessary to install software that is not (yet)
available for a given distribution.

To address this concern, one must archive a binary copy of the image
used at the time the research was first performed.  Docker provides a
simple utility to save an image as a portable `tarball` file that can
be read in by any other Docker installation, providing a robust way to
run the exact versions of all software involved.  By testing both the
`tarball` archive and the image generated by the latest Dockerfile,
Docker provides a simple way to confirm whether or not code rot has
effected the function of a particular piece of code.

<!-- FIXME move this later --> 
To simplify this process, Docker also supports *Automated Builds*
through the Docker Hub.  This acts as a kind of Continous Integration
(CI) service that verifies the image builds correctly whenever the
Dockerfile is updated, particularly if the Dockerfile includes checks
for the environment. The Hub also provides a convenient distribution
service, freely storing the prebuilt images, along with their metadata,
for download and reuse by others. The Docker Hub is a free service and
an open source software product so that users can run their own private
versions of the Hub on their own servers, for instance, if security of
the data or the longevity of the public platform is a concern.

### 4. Barriers to adoption and re-use

A technical solution, no matter how elegant, will be of little practical 
use for reproducible research unless it is both easy to use and adapt to
the existing workflow patterns of practicing domain researchers.

Though most of the concerns we have discussed so far can be addressed
through well-designed workflow software or the use of a DevOpts approach
to provisioning virtual machines by scripts, neither approach has seen
widespread adoption by domain researchers, who work primarily in a
local rather than cloud-based environment using develoment tools native
to their personal operating system. To gain more widespread adoption,
reproducible research technologies must make it easier, not harder,
for a researcher to perform the tasks they are already doing (before 
considering any additional added benefits).

These issues are reflected both during the original research or development
phase and in any subsequent reuse. Another researcher may be less likely
to build on existing work if it can only be done by using a particular
workflow system or monolothic software platform with which they are
unfamiliar.  Likewise, a user is more likey to make their own computational
environment avialable for reuse if it does not involve a significant
added effort in packaging and documenting [Stodden 2010].

Though Docker is not immune to these challenges, it offers a interesting 
example of a way forward in addressing these fundamental cocerns.

- *Local environment*
- *Modular reuse*

<!---
- The ability to work in a familiar, local environment (during one's own research or when re-using elements from another researcher)
- An Docker container approach to both development and deployment
- Image re-use & Image linking 
-->

# Using Docker as a local development environment

@@

- Shared volumes
- Host OS tools

@@

In the R environment, the open source RStudio IDE provides
another key component in making this system accessible to most
domain scientists. Because the RStudio IDE is written completely with
standard web languages (javascript, css, and html), its web-based 
RStudio Server looks and feels identical to its popular desktop IDE.
RStudio Server provides a way to interact with R on a remote environment
without the latency, X tunnelling, etc typically involved in running R
on a remote server.

@@

## Docker remix: Building re-usable modules

@@

----------------

- Interactive use, scripted use (Dockerfiles)
- Reverting to a previous state
- Integration with local development environment
- Integration with cloud computing
- Transparent & extensible
- serial and linking containers
- Best practices are still taking shape

----------

##  Best practices

- Using dockerfiles instead of containers
- Using appropriate, common base images
- Archiving tarball snapshots
- Adding tests/checks to the Dockerfile
- Sharing 
- Leveraging package managers appropriately to handle software dependencies

Upcoming features:

- signed containers?

-------------

@@

@@

--------------------------------

<!-- Covered elsewhere in a more integrated fashion -->
Researchers familiar with LXC may be tempted to view the Docker
system as little more than a creative application and marketing of an
existing technology. However, the Docker approach adds several features
crucial for its use in reproducible research (adapted from [Docker
FAQ](https://docs.docker.com/faq/#what-does-docker-add-to-just-plain-lxc)):
